---
title: "U.S. Severe Weather Impact Analysis"
author: "Aaron Cole"
date: "19 January 2016"
output: 
    html_document:
      number_sections: yes
      theme: united
      toc: yes
---

# Synopsis #
* The analysis explores two aspects of the impact of severe weather events in the U.S.
** Which event types are most harmful with respect to population health.
** Which event types have the greatest economic consequences.
* The analysis will use historic data to demonstrate which event types are the most impactful, using data from the Storm Events Database maintained by the U.S. National Oceanic and Atmospheric Administration.
* Data for events beginning prior to 1st January 1996 has been excluded, as before this point only a small number of specific event types were being tracked.
* The remaining data comprises of over 650K observations from 1st January 1996 onwards, for all severe weather event types.
* Our anlysis will focus on the variables for fatality and injury counts, as well as property and crop damage in USD.

# Data processing #

## Load libraries ##
```{r echo=TRUE, cache=TRUE}
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Download the data ##
```{r echo=TRUE, cache=TRUE}
# Check if the file already exists locally before downloading.
if(file.exists("data.bz2") == FALSE) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  "data.bz2",
                  mode = "wb")
}
```

## Subset by year and remove variables that are not required ##
Refer to the [NOAA Storm Events Database Details](http://www.ncdc.noaa.gov/stormevents/details.jsp) webpage for information about the event types being monitored by year. Monitoring for all events started in January 1996 and continues to the present day. Our analysis is interested in examining all events and understanding which event types are most harmful to population health and which event types have the greatest economic consequences. We will exclude data for all years before 1996. The resulting subsetted data represents monitoring of all event types for 1996 onwards and still provides a sizable sample across many years.
```{r echo=TRUE, cache=TRUE}
# Read the data archive.
weatherData <- tbl_df(read.csv("data.bz2"))

# Convert the BGN_DATE variable from factor to date.
weatherData$BGN_DATE <- as.Date(weatherData$BGN_DATE, format = "%m/%d/%Y")

# Filter out observations where variable BGN_DATE is before 1996.
weatherData <- filter(weatherData, BGN_DATE %within% interval(ymd(19960101), Sys.Date()))

# Select only the required variables.
weatherData <- weatherData %>%
select(event = EVTYPE,
       fatalities = FATALITIES,
       injuries = INJURIES,
       property_damage = PROPDMG,
       property_damage_function = PROPDMGEXP,
       crop_damage = CROPDMG,
       crop_damage_function = CROPDMGEXP)
```

## Sort out the property damage value ##
```{r echo=TRUE, cache=TRUE}
# Convert the property damage function column to character.
weatherData$property_damage_function <- as.character(weatherData$property_damage_function)

# Convert the values ready for multiplication.
weatherData$property_damage_function[weatherData$property_damage_function == ""] <- "1"
weatherData$property_damage_function[weatherData$property_damage_function == "0"] <- "1"
weatherData$property_damage_function[weatherData$property_damage_function == "K"] <- "1000"
weatherData$property_damage_function[weatherData$property_damage_function == "M"] <- "1000000"
weatherData$property_damage_function[weatherData$property_damage_function == "B"] <- "1000000000"

# Convert to integer.
weatherData$property_damage_function <- as.integer(weatherData$property_damage_function)

# Calculate the new column and drop the old ones.
weatherData <- weatherData %>%
    mutate(property_damage_usd = property_damage * property_damage_function) %>%
    select(-property_damage, -property_damage_function)
```

## Sort out the crop damage value ##
```{r echo=TRUE, cache=TRUE}
# Convert the crop damage function column to character.
weatherData$crop_damage_function <- as.character(weatherData$crop_damage_function)

# Convert the values ready for multiplication.
weatherData$crop_damage_function[weatherData$crop_damage_function == ""] <- "1"
weatherData$crop_damage_function[weatherData$crop_damage_function == "0"] <- "1"
weatherData$crop_damage_function[weatherData$crop_damage_function == "K"] <- "1000"
weatherData$crop_damage_function[weatherData$crop_damage_function == "M"] <- "1000000"
weatherData$crop_damage_function[weatherData$crop_damage_function == "B"] <- "1000000000"

# Convert to integer.
weatherData$crop_damage_function <- as.integer(weatherData$crop_damage_function)

# Calculate the new column and drop the old ones.
weatherData <- weatherData %>%
    mutate(crop_damage_usd = crop_damage * crop_damage_function) %>%
    select(-crop_damage, -crop_damage_function)
```

## Combine the property and crop damage values into one column ##
```{r echo=TRUE, cache=TRUE}
# Calculate the new column and drop the old ones.
weatherData <- weatherData %>%
    mutate(damage_usd = property_damage_usd + crop_damage_usd) %>%
    select(-property_damage_usd, -crop_damage_usd)
```

## Prepare event economic impact data ##
```{r echo=TRUE, cache=TRUE}
# Select required columns and filter out zero values.
economicData <- weatherData %>% 
    select(event, damage_usd) %>%
    filter(damage_usd > 0)

# Group by event and sum the damages.
economicData <- economicData %>% 
    group_by(event) %>%
    summarise_each(funs(sum))
```

## Prepare event health impact data ##

# Results #

## Events with the greatest economic consequences ##

### Add new column for event damages as a percentage of total ###
Add a new percent column to show the proportion of the damages for each event, as a percent of the total for all events.
```{r echo=TRUE, cache=TRUE}
economicData <- mutate(economicData, percent = round((damage_usd / sum(economicData$damage_usd) * 100), 1))
```
Examine the data with the quantile function.
```{r echo=TRUE, cache=TRUE}
quantile(economicData$percent, seq(0, 1, 0.05))
```

### Reducing the number of events ###
We can see that 85% of our events have a damage percentage less than or equal to 0.025%. Subset for these events and check the total sum of the damages for the subset to verify if they can be removed without significantly impacting the total damages for the complete dataset.
```{r echo=TRUE, cache=TRUE}
economicData[economicData$percent <= 0.025,]
round((sum(economicData[economicData$percent <= 0.025,]$damage_usd) / sum(economicData$damage_usd)) * 100, 3)
```
The 158 events to be removed only represent a damage value of 0.335% of the total damages for all events, removing these 159 event types should not significantly affect our analysis.
```{r echo=TRUE, cache=TRUE}
economicData <- economicData[economicData$percent > 0.025,]
```

### Tidy up of the variables ###
```{r echo=TRUE, cache=TRUE}
# Drop the percent variable.
economicData <- select(economicData, -percent)

# Convert the event variable to character.
economicData$event <- as.character(economicData$event)

# Tidy up the events into sensible groupings.
economicData$event[economicData$event %in% c("DROUGHT", 
                                             "EXCESSIVE HEAT")] <- "DROUGHT/HEAT"

economicData$event[economicData$event %in% c("EXTREME COLD", 
                                             "FROST/FREEZE")] <- "FREEZE/COLD"

economicData$event[economicData$event %in% c("HAIL")] <- "HAIL"

economicData$event[economicData$event %in% c("HEAVY SNOW", 
                                             "WINTER STORM", 
                                             "BLIZZARD", 
                                             "ICE STORM")] <- "SNOW/ICE"

economicData$event[economicData$event %in% c("LANDSLIDE")] <- "LANDSLIDE"

economicData$event[economicData$event %in% c("HEAVY RAIN", 
                                             "STORM SURGE", 
                                             "STORM SURGE/TIDE")] <- "RAIN/SURGE"

economicData$event[economicData$event %in% c("HIGH WIND", 
                                             "STRONG WIND", 
                                             "THUNDERSTORM WIND", 
                                             "TSTM WIND", 
                                             "TROPICAL STORM", 
                                             "TORNADO")] <- "WIND/STORM/TORNADO"

economicData$event[economicData$event %in% c("WILD/FOREST FIRE", 
                                             "WILDFIRE")] <- "FIRE"

economicData$event[economicData$event %in% c("COASTAL FLOOD", 
                                             "FLOOD", 
                                             "FLASH FLOOD")] <- "FLOOD"

economicData$event[economicData$event %in% c("HURRICANE", 
                                             "HURRICANE/TYPHOON", 
                                             "TYPHOON")] <- "HURRICANE/TYPHOON"

economicData$event[economicData$event %in% c("LIGHTNING")] <- "LIGHTNING"

# Group on event and sum the damage value.
economicData <- economicData %>% 
    group_by(event) %>%
    summarise_each(funs(sum))

# Add a new variable for the damage in billions of USD for easier plotting.
economicData <- mutate(economicData, damage_usd_billions = round(damage_usd / 1000000000, 2))
```

### Plot the economic impact data ###
```{r echo=TRUE, cache=TRUE}
# Create the plot.
p <- ggplot(data = economicData, aes(x = event, y = damage_usd_billions)) +
    geom_bar(stat="identity") +
    xlab("Event") +
    ylab("Damage in billions of USD") +
    ggtitle("Damage in Billions of USD by Event") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

print(p)
```

### Print summary table ###
```{r echo=TRUE, cache=TRUE}
economicData <- economicData %>%
    mutate(percentage = round((damage_usd / sum(economicData$damage_usd) * 100), 1)) %>%
    select(-damage_usd) %>%

print(economicData)
```

### Summary of event economic impact ###
The top four event types by damage value, account for 87% of the total damage value across all events.

Event Type | Damage as % of Total | Damage USD | Notes
-----------|----------------------|------------|------
FLOOD | 41% | $165B |
HURRICANE/TYPHOON | 22% | $87B |
RAIN/SURGE | 12% | $49B |
WIND/STORM/TORNADO | 12% | $48B |
OTHERS | 13% | $51B | Covers the remaining 7 event types.